---
title: "Basis Expansions and the bex Class"
author: "McNeil, Neslehova, Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basis Expansions and the bex Class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 4,
  fig.align = "center"
)
```

```{r library}
library(basiscor)
library(udp)
```

`basisexpand()` looks for the pair of functions `g`, `h` -- each a linear
combination of the first `maxorder` orthonormal basis functions (shifted
Legendre polynomials, or a cosine basis) -- that maximizes the generalized
Spearman correlation between `g(X1)` and `h(X2)` for a bivariate sample `X`.
The result is a \linkS4class{bex} object, which also carries the pair of
**udp** transformations `Tg(u) = F(g(u))`, `Th(u) = F(h(u))` built from `g`
and `h`. This vignette illustrates `bex` on the same awkwardly-dependent
dataset that the **udp** package's own vignette (`vignette("udp", package =
"udp")`) uses to illustrate `aceshuffle()`, so the two can be compared
directly: `aceshuffle()` looks for the best pair of piecewise-linear
*shuffles*, `basisexpand()` for the best pair of smooth, low-dimensional
*polynomial* (or cosine) transformations.

## The same folded data as the udp vignette

Two degree-fixed **udp** transformations -- `udplegendre(5)` and
`udpcosine(6)` -- are used to fold each margin of a strongly dependent Gumbel
copula sample, via stochastic inversion (`udpsi()`). This destroys the
ordinary correlation almost entirely while leaving a highly structured,
non-monotonic dependence behind.

```{r data}
T5 <- udplegendre(5)
T3 <- udpcosine(6)

set.seed(1)
n <- 1000
GC <- copula::gumbelCopula(3)
W <- copula::rCopula(n, GC)
X <- cbind(udpsi(T5, W[, 1]), udpsi(T3, W[, 2]))

cor(W)[1, 2]
cor(X)[1, 2]
```

```{r data-plot}
plot(X, pch = 20, cex = 0.4, xlab = "X1", ylab = "X2")
```

## Finding a basis expansion

`basisexpand()` computes the `maxorder` by `maxorder` matrix of sample basis
correlations between the two margins (via `basiscormatrix()`) and takes its
leading singular vectors as the weights `alphag`, `alphah` of `g` and `h`.

```{r basisexpand}
bex <- basisexpand(X)
bex
```

`maxcor` is the generalized Spearman correlation attained by this pair --
already well above the raw `cor(X)`, and, as the next section shows, close to
the correlation of the original copula sample `W`.

The raw expansions `g`, `h` can be plotted side by side:

```{r plot-raw}
plot(bex)
```

## The recovered udp transformations

With `udp = TRUE`, `plot()` draws the udp transformations `Tg`, `Th` instead
of the raw expansions, again side by side.

```{r plot-udp}
plot(bex, udp = TRUE)
```

Overlaying each fitted transformation against the *true* one used to build
its margin of `X` checks the recovery directly: `Tg` against `T5` (margin 1),
`Th` against `T3` (margin 2).

```{r plot-udp-overlay, fig.width = 8}
u <- seq(0, 1, length.out = 400)
op <- par(mfrow = c(1, 2))
plot(u, udptrans(bex@Tg, u), type = "l", xlab = "u", ylab = "T(u)", asp = 1, main = "margin 1")
lines(u, udptrans(T5, u), col = "red", lty = 2)
legend("bottomright", legend = c("fitted Tg", "true T5"), col = c("black", "red"), lty = c(1, 2), cex = 0.8)

plot(u, udptrans(bex@Th, u), type = "l", xlab = "u", ylab = "T(u)", asp = 1, main = "margin 2")
lines(u, udptrans(T3, u), col = "red", lty = 2)
legend("bottomright", legend = c("fitted Th", "true T3"), col = c("black", "red"), lty = c(1, 2), cex = 0.8)
par(op)
```

Both curves are highly oscillatory, which makes an exact-looking overlay by
eye hard even when the recovery is good; a more direct check is the
correlation between the fitted and true transformations evaluated on a grid,
which is high for the matching pair and near zero for the mismatched one:

```{r plot-udp-check}
data.frame(
  cor_Tg_T5 = cor(udptrans(bex@Tg, u), udptrans(T5, u)),
  cor_Tg_T3 = cor(udptrans(bex@Tg, u), udptrans(T3, u)),
  cor_Th_T5 = cor(udptrans(bex@Th, u), udptrans(T5, u)),
  cor_Th_T3 = cor(udptrans(bex@Th, u), udptrans(T3, u))
)
```

As with `aceshuffle()`'s shuffles (see the **udp** vignette), the fitted pair
is unique only up to the joint reflection `(g, h) -> (-g, -h)`, equivalently
`(Tg, Th) -> (1 - Tg, 1 - Th)`, which leaves the correlation unchanged;
`basisexpand()` picks an orientation but it is not guaranteed to match the one
used to generate the data. If an overlay comes out reflected, compare against
`1 - udptrans(...)` instead.

## Undoing the fold

`udptrans()` applied to a \linkS4class{bex} object transforms both columns of
a bivariate `(0, 1)` matrix at once, `cbind(udptrans(Tg, X[, 1]), udptrans(Th,
X[, 2]))`, recovering data that is once again close to linearly correlated:

```{r udptrans-bex}
V <- udptrans(bex, X)
plot(V, pch = 20, cex = 0.4, xlab = "V1", ylab = "V2")
c(fitted = bex@maxcor, achieved = cor(V)[1, 2], original = cor(W)[1, 2])
```

## References

- McNeil, A. J. (2021). Modelling volatile time series with v-transforms and
  copulas. *Risks*, **9**(1), 14.
- McNeil, A. J., Nešlehová, J. G. and Smith, A. D. (2025). Measures and models
  of non-monotonic dependence. arXiv:2512.10828.
