---
title: "Orthonormal Basis Correlations of Various Copulas"
author: "Alexander J. McNeil"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Orthonormal Basis Correlations of Various Copulas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(copula)
library(rvinecopulib)
library(basiscor)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`basiscor(object, j, k)` computes the population basis correlation between
the degree-`j` and degree-`k` orthonormal basis functions (shifted Legendre
polynomials by default, or a cosine basis with `type = "cosine"`) of the two
margins of a copula. At degree `j = k = 1`, this is exactly Spearman's rho:
the orthonormal degree-1 shifted Legendre polynomial is `P_1(u) = sqrt(3)(2u
- 1)`, so `basiscor(cop, 1, 1) = 3 * E[(2U-1)(2V-1)] = 12 Cov(U, V)`, which is
the usual definition of Spearman's rho. This vignette checks `basiscor()`'s
population values against the **copula** package's own `rho()` where
available, and demonstrates the rest of the package's calculations --
`basiscordata()` for the sample analogue, `basiscormatrix()` for a full
matrix of basis correlations, and `extremalLegendre()` for the bounds
`basiscor()` can attain.

`basiscor()` has two ways of computing a population value from a copula
object: `method = "p"` integrates against the copula's distribution function
`pCopula()` (the default -- faster, and the only option when a family has no
closed-form density), and `method = "d"` integrates against the density
`dCopula()` (needed when a family has no closed-form distribution function,
such as the `t` copula below). The two should agree closely whenever both are
available, which is itself a useful consistency check.

### FGM and Frank Copulas

These have simple closed-form densities and distribution functions, and
`basiscor()` agrees with `rho()` to numerical precision.

```{r}
cop <- fgmCopula(1)
rho(cop)
basiscor(cop)
basiscor(cop, 1, 2)
basiscor(cop, 2, 2)

cop <- frankCopula(2)
rho(cop)
basiscor(cop)
basiscor(cop, 1, 2)
basiscor(cop, 2, 2)
```

### Gumbel, Clayton and Joe Copulas (with tail dependence)

`method = "p"` and `method = "d"` agree closely with each other here, as they
should, but `basiscor(cop)` (i.e. `j = k = 1`, so this should equal Spearman's
rho) does not match `copula::rho()` as closely as in the FGM/Frank case above
-- a few times `1e-3` for Clayton, a few times `1e-4` for Gumbel. Spearman's
rho is not implemented for the Joe copula in the **copula** package at all.

```{r}
cop <- claytonCopula(1.5)
rho(cop)
basiscor(cop)
basiscor(cop, method = "d")
basiscor(cop, 1, 2)
basiscor(cop, 2, 2)

cop <- gumbelCopula(1.5)
rho(cop)
basiscor(cop)
basiscor(cop, method = "d")
basiscor(cop, 1, 2)
basiscor(cop, 2, 2)

cop <- joeCopula(1.5)
basiscor(cop)
basiscor(cop, method = "d")
basiscor(cop, 1, 2)
basiscor(cop, 2, 2)
```

A direct large-sample check settles which side the discrepancy is on: a
Monte Carlo estimate of Spearman's rho from a very large simulated sample
lands close to `basiscor()`'s value in both cases, and well outside the
range `copula::rho()`'s value would suggest given the Monte Carlo standard
error. `copula::rho()` for these two families appears to be the less
accurate of the two, not `basiscor()`.

```{r rho-montecarlo}
set.seed(1)
n <- 5e6
for (cop in list(claytonCopula(1.5), gumbelCopula(1.5))) {
  U <- rCopula(n, cop)
  rho_mc <- cor(U[, 1], U[, 2], method = "spearman")
  cat(class(cop)[1], "\n")
  cat("  copula::rho()   =", rho(cop), "\n")
  cat("  basiscor()      =", basiscor(cop), "\n")
  cat("  Monte Carlo     =", rho_mc, "(se ~", sqrt((1 - rho_mc^2) / n) * 1.5, ")\n\n")
}
```

### Gauss Copula

This case gives some confidence in the integrals: `method = "p"` matches the
exact closed form and `copula::rho()` to the displayed precision, and
`method = "d"` is only out at the 7th digit after the point.

```{r}
rho <- 0.7
cop <- normalCopula(rho)
(6 / pi) * asin(rho / 2)
rho(cop)
basiscor(cop)
basiscor(cop, method = "d")
basiscor(cop, 1, 2)
basiscor(cop, 2, 2)
```

### t Copula

Spearman's rho is not implemented for the `t` copula in the **copula**
package. Its distribution function is also only available for integer
degrees of freedom, which is why `basiscor()` defaults to `method = "d"`
automatically whenever `df` is not an integer.

```{r}
rho <- 0.7
nu <- 4
cop <- tCopula(rho, df = nu)
basiscor(cop)
basiscor(cop, method = "d")
basiscor(cop, 1, 2)
basiscor(cop, 2, 2)
```

### Copulas from rvinecopulib

`basiscor()` also accepts a `bicop_dist` bivariate copula object from the
**rvinecopulib** package, dispatching to `basiscorcopula(cop, j, k, copobj =
"bicop")`, which integrates against `rvinecopulib::pbicop()` (`method =
"p"`) or `rvinecopulib::dbicop()` (`method = "d"`). The population values
agree with the equivalent **copula** package object to numerical precision.

```{r}
cop <- claytonCopula(1.5)
bc <- bicop_dist("clayton", 0, 1.5)
basiscor(cop, 1, 1)
basiscor(bc, 1, 1)
basiscor(cop, 2, 3)
basiscor(bc, 2, 3)
```

Unlike **copula**'s `tCopula`, `rvinecopulib`'s `bicop_dist` `t` copula has a
distribution function that works for any degrees of freedom, not just
integer ones, so `method = "p"` needs no automatic fallback to `method =
"d"` the way `basiscorcopula()` does for `copobj = TRUE`. The two methods
still agree closely, as they should:

```{r}
bc <- bicop_dist("t", 0, c(0.5, 4.3))
basiscor(bc, 1, 1, method = "p")
basiscor(bc, 1, 1, method = "d")
```

## Sample basis correlations

`basiscordata()` computes the sample analogue of `basiscor()` from a bivariate
data matrix, using only the ranks of the two columns -- it is a
generalization of Spearman's rho in the same sense as `basiscor()` is at the
population level. At `j = k = 1` with the default `method = "T3"`, it is
*exactly* the ordinary sample Spearman correlation.

```{r}
set.seed(1)
cop <- claytonCopula(1.5)
X <- rCopula(2000, cop)

basiscor(cop, 1, 1) # population value
basiscordata(X, 1, 1) # sample value
cor(X[, 1], X[, 2], method = "spearman") # identical to basiscordata(X, 1, 1)
```

`basiscordata()` offers several other estimators (`method = "T1"` to
`"T6"`), differing in how the ranks are turned into pseudo-observations in
`(0, 1)` and whether the result is a covariance- or correlation-type
statistic; see `?basiscordata`.

## A matrix of basis correlations

`basiscormatrix()` computes `basiscor()` (or `basiscordata()`, for a data
matrix) for every pair of degrees up to `maxorder` at once.

```{r}
basiscormatrix(cop, maxorder = 4)
```

## Extremal basis correlations

For the shifted-Legendre basis, `extremalLegendre(j, k)` gives the maximum
(or, with `case = "min"`, the minimum) value `basiscor(cop, j, k)` can attain
over *any* copula -- so every population basis correlation should lie between
the two.

```{r}
basiscor(cop, 3, 4)
c(min = extremalLegendre(3, 4, "min"), max = extremalLegendre(3, 4, "max"))
```
